### 2022-07-09 pre/pre.tex
geoff ch 3

We distinguish a set of base types $T$ which is required to have the Boolean
type $o$, and a set of symbols $\Sigma$. To each symbol $\cst{f} \in \Sigma$ a
tuple $(\tau_1, \ldots, \tau_{n}, \tau), n \geq 0$ of types is assigned, written
$\cst{f} : (\tau_1, \ldots, \tau_{n}) \rightarrow \tau$. We say that $(\tau_1,
\ldots, \tau_{n})$ are \emph{argument types}, $\tau$ is the \emph{return type}, and $n$
is the \emph{arity} of the symbol $\cst{f}$. If the return type of $\cst{f}$ is $o$, we call
$\cst{f}$ a \emph{predicate symbol}, otherwise we call it a \emph{function symbol}. Argument types may not be Boolean.

@@

\subsection{Term Order and Selection}
\label{sec:pre:order}
Superposition calculus is parameterized by a
term order $\succ$ with the following properties:

\noindent\begin{tabular}{p{\dimexpr 0.3\linewidth-2\tabcolsep}p{\dimexpr 0.7\linewidth-2\tabcolsep}}
    \textit{Irreflexive} & For all terms $s$, $s \not\succ s$ \\
    \textit{Transitive} & For all terms $s, t, u$,  if $s\succ t$ and $t \succ u$ then $s \succ u$ \\ 
    \textit{Subterm property} & For all terms $s$ and contexts $C$, $C[s] \succ s$ \\
    \textit{Respects substitutions} & For all terms $s, t$ and substitutions $\sigma$, $s \succ t$ implies $\sigmaterm{s} \succ \sigmaterm{t}$ \\
    \textit{Respects contexts} & For all terms $s,t$ and nonempty contexts $C$, $s \succ t$ implies $C[s] \succ C[t]$ \\
    \textit{Ground total} & For all distinct
    ground terms $s$ and $t$ either $s \succ t$ or $t \succ s$ \\
    \textit{Well-founded} & There are no infinite chains of the form $s_1 \succ s_2 \succ \cdots$ 
\end{tabular}

@@

Calculi from the resolution family, including superposition, perform inferences
only on unifiable terms. We say terms $s$ and $t$ are {\em unifiable} if there is a
substitution $\sigma$ such that $\sigmaterm{s} = \sigmaterm{t}$; we further say
$\sigma$ is a unifier. A \emph{unification constraint} $s \unif t$ is a pair of two terms of the same type. 
We always specify if the constraint should be interpreted as ordered or unordered.
We say a substitution $\sigma$ is \emph{more general} than $\varrho$ if there exists a substitution $\theta$
such that $\varrho = \theta\sigma$.
A \emph{most general unifier} ({MGU}) is a
unifier $\sigma$, such that for any other unifier $\varrho$, $\sigma$ is more general than $\varrho$. In first-order logic,
the MGU is unique up to variable renaming. For example, $\cst{f}(X) \unif \cst{f}(Y)$ yields the MGU 
$\{ X \mapsto Y\}$ or  $\{ Y \mapsto X \}$. First-order logic also admits efficient algorithms for computing MGUs \cite{hv-09-unifalgs}.

@@

The rules described above are used to generate new clauses from already derived
ones. Even though orders and selection are used to reduce the search space,
it still grows rather fast. To identify unnecessary clauses in the search
space, superposition features a \emph{redundancy criterion}
\cite[Sect.~4.2.2]{bg-01-resolution}. We say that a ground clause $C$ is
{\em redundant} in a ground set of clauses $N$ if it is entailed by a subset of
clauses in $N$ such that each clause in this subset is $\succ$-smaller than $C$. More
generally, a clause (ground or not) is redundant in set $N$ if every ground
instance of $C$ is redundant for the grounding of $N$. Removing redundant
clauses does not affect completeness of superposition.

Next to the four generating rules, superposition provers also use
simplification rules, which are justified by the redundancy criterion. These
rules effectively replace the pre\-mises with conclusions, and we denote them
using double bars. In theory, the rules work by adding conclusions and then
using them to show that premises are redundant; after this, premises can be eliminated. As
an example of such a rule, let us introduce rewriting (demodulation) of negative
literals (\infname{RN}):

@@

\looseness=-1
In Chapter \ref{ch:intro} we described a roadmap (consisting of three steps) to
extend an efficient superposition prover to higher-order logic. Together with
our colleagues we did the necessary theoretical work by extending the
superposition calculus using the same roadmap:

\begin{enumerate}
    \item Bentkamp, Blanchette, Cruanes, and Waldmann designed a complete
    superposition calculus for higher-order logic devoid of
    $\lambda$-abstraction and first-class Booleans \cite{bbcw-21-lfho}. This calculus is called the \emph{Boolean-free $\lambda$-free higher-order superposition calculus} 
    and we shortly refer to it as \lfsup{}.
    \item \begin{sloppypar}
    Bentkamp, Blanchette, Tourret,  Vukmirović, and Waldmann
    extended this calculus to support $\lambda$-abstraction
    \cite{bbtvw-21-sup-lam}. They call this new calculus the \emph{Boolean-free $\lambda$-superposition calculus},
    and we shortly refer to it as \lsup{}.
    \end{sloppypar}        
    \item Bentkamp, Blanchette, Tourret, and Vukmirovi\'c extended \lsup{} with
    support for first-class Booleans \cite{bbtv-21-full-ho-sup}. This amounts to
    designing the complete calculus for full higher-order logic. This
    calculus is called Boolean $\lambda$-superposition, or shortly \osup{}.
\end{enumerate}
%
Note that we use the name $\lambda$-superposition for both \lsup{}
and \osup{}.

@@

\ourpara{E} E is a state-of-the-art first-order prover based on superposition.
In the last decade, together with its derivatives, it has revolved around the
second place at the first-order division of the CASC theorem prover
competition \cite{ss-96-casc}. It is open-source, written in C (without using any
external libraries), and designed around efficient algorithms and data structures
rather than a highly optimized codebase. This makes it a good target for implementing
successful higher-order techniques previously evaluated in Zipperposition.

>>>>

We distinguish a set of base types $T$ which is required to include the Boolean
type $o$, and a set of symbols $\Sigma$. To each symbol $\cst{f} \in \Sigma$ a
tuple $(\tau_1, \ldots, \tau_{n}, \tau), n \geq 0$ of types is assigned, written
$\cst{f} : (\tau_1, \ldots, \tau_{n}) \rightarrow \tau$. We say that $(\tau_1,
\ldots, \tau_{n})$ are \emph{argument types}, $\tau$ is the \emph{return type}, and $n$
is the \emph{arity} of the symbol $\cst{f}$. If the return type of $\cst{f}$ is $o$, we call
$\cst{f}$ a \emph{predicate symbol}, otherwise we call it a \emph{function symbol}. Argument types may not be Boolean.

@@

\subsection{Term Order and Selection}
\label{sec:pre:order}
Superposition calculus is parameterized by a
term order $\succ$ with the following properties:
%
\noindent\begin{tabular}{p{\dimexpr 0.3\linewidth-2\tabcolsep}p{\dimexpr 0.7\linewidth-2\tabcolsep}}
    \textit{Irreflexive} & For all terms $s$, $s \not\succ s$ \\
    \textit{Transitive} & For all terms $s, t, u$,  if $s\succ t$ and $t \succ u$ then $s \succ u$ \\ 
    \textit{Subterm property} & For all terms $s$ and contexts $C$, $C[s] \succ s$ \\
    \textit{Respects substitutions} & For all terms $s, t$ and substitutions $\sigma$, $s \succ t$ implies $\sigmaterm{s} \succ \sigmaterm{t}$ \\
    \textit{Respects contexts} & For all terms $s,t$ and nonempty contexts $C$, $s \succ t$ implies $C[s] \succ C[t]$ \\
    \textit{Ground total} & For all distinct
    ground terms $s$ and $t$ either $s \succ t$ or $t \succ s$ \\
    \textit{Well-founded} & There are no infinite chains of the form $s_1 \succ s_2 \succ \cdots$ 
\end{tabular}

@@

Calculi from the resolution family, including superposition, perform inferences
only on unifiable terms. We say terms $s$ and $t$ are {\em unifiable} if there is a
substitution $\sigma$ such that $\sigmaterm{s} = \sigmaterm{t}$; we further say
$\sigma$ is a unifier. A \emph{unification constraint} $s \unif t$ is a pair of two terms of the same type. 
We always specify if the constraint should be interpreted as ordered or unordered.
We say a substitution $\sigma$ is \emph{more general} than $\varrho$ if there exists a substitution $\theta$
such that $\varrho = \theta\sigma$.
A \emph{most general unifier} ({MGU}) is a
unifier $\sigma$, such that for any other unifier $\varrho$, $\sigma$ is more general than $\varrho$. In first-order logic,
the MGU is unique up to variable renaming. For example, $\cst{f}(X) \unif \cst{f}(Y)$ yields the MGU 
$\{ X \mapsto Y\}$ or  $\{ Y \mapsto X \}$. First-order logic also admits efficient, linear-time algorithms for computing MGUs \cite{hv-09-unifalgs}.

@@

The rules described above are used to generate new clauses from already derived
ones. Even though orders and selection are used to reduce the search space,
many unnecessary clauses might be created. To identify unnecessary clauses in the search
space, superposition features a \emph{redundancy criterion}
\cite[Sect.~4.2.2]{bg-01-resolution}. We say that a ground clause $C$ is
{\em redundant} in a ground set of clauses $N$ if it is entailed by a subset of
clauses in $N$ such that each clause in this subset is $\succ$-smaller than $C$. More
generally, a clause (ground or not) is redundant in set $N$ if every ground
instance of $C$ is redundant for the grounding of $N$. Removing redundant
clauses does not affect completeness of superposition.

Next to the four generating rules, superposition provers also use
simplification rules, which are justified by the redundancy criterion. These
rules replace the pre\-mises with conclusions, and we denote them
using double bars. In theory, the rules work by adding conclusions and then
using them to show that premises are redundant; after this, premises can be eliminated. As
an example of such a rule, let us introduce rewriting (demodulation) of negative
literals (\infname{RN}):

@@

\looseness=-1
In Chapter \ref{ch:intro} we described a roadmap (consisting of three steps) to
extend an efficient superposition prover to higher-order logic. My colleagues
and me did the necessary theoretical work by extending the superposition calculus
using the same roadmap:

\begin{enumerate}
    \item Bentkamp, Blanchette, Cruanes, and Waldmann designed a complete
    superposition calculus for higher-order logic devoid of
    $\lambda$-abstraction and first-class Booleans \cite{bbcw-21-lfho}. This calculus is called the \emph{Boolean-free $\lambda$-free higher-order superposition calculus} 
    and we shortly refer to it as \lfsup{}.
    \item \begin{sloppypar}
    Bentkamp, Blanchette, Tourret,  Vukmirović, and Waldmann
    extended this calculus to support $\lambda$-abstraction
    \cite{bbtvw-21-sup-lam}. They call this new calculus the \emph{Boolean-free $\lambda$-superposition calculus},
    and we shortly refer to it as \lsup{}.
    \end{sloppypar}        
    \item Bentkamp, Blanchette, Tourret, and Vukmirovi\'c extended \lsup{} with
    support for first-class Booleans \cite{bbtv-21-full-ho-sup}. This amounts to
    designing the complete calculus for full higher-order logic. This
    calculus is called Boolean $\lambda$-superposition, or shortly \osup{}.
\end{enumerate}
%
Note that in this thesis the name $\lambda$-superposition is used for both \lsup{}
and \osup{}.

@@

\ourpara{E} E is a state-of-the-art first-order prover based on superposition.
In the last decade, together with its derivatives, it was usually the
second place at the first-order division of the CASC theorem prover
competition \cite{ss-96-casc}. It is open-source, written in C (without using any
external libraries), and designed around efficient algorithms and data structures
rather than a highly optimized codebase. This makes it a good target for implementing
successful higher-order techniques previously evaluated in Zipperposition.
